---
title: "Rates of Contraceptive Use For Women of Different Age Groups"
author: "Megan Tran"
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
---







<br> 







 Link to the pdf report here:





<br>






# Background on the Data and Study Sample

```{r setup, message=FALSE, echo=FALSE, warning=FALSE}

library(lubridate)
library(tidyverse)
library(data.table)
library(ggplot2)
library(dtplyr)
library(dplyr)
library(stringr)
library(knitr)
library(DT)

# Initialize code chunk options
opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  eval=TRUE,
  echo = TRUE,
  cache = FALSE,
  fig.width = 7, 
  fig.align = 'center',
  fig.asp = 0.618,
  out.width = "700px")
```


```{css, echo = FALSE}
.code-r { /* Code block */
  font-size: 15px;
}

.code-r-small { /* Code block */
  font-size: 10px;
}
```



The dataset used for this research report was downloaded from the California Health and Human Services Agency (CHHS) Open Data Portal: https://data.chhs.ca.gov/dataset/contraceptive-care-for-women-use-by-race-ethnicity-contraceptive-type-and-age-group/resource/12a73f54-dcf4-4e38-843c-e988385be69b.
The Contraceptive Care - All Women measure (CCW), as part of the Maternal and Infant Health Initiative, Contraceptive Care Quality grant, was compiled data taken from eligible women ages 15-44 who are at risk for unintended pregnancy. It reported the rates of women's contraceptive use broken down by year (2014-2016), age group (15-20 vs. 21-44), and race. Contraceptive type either fell under the category of long-acting reversible methods of contraception (LARC), such as birth control implants and IUDs, or most/moderately effective methods of contraceptions (M/M), which include oral birth control pills, patches, rings, injectables, female sterilization, and diaphragms. 
 
 
 
 
 
 <br>
 
 
 
 
 
 
# Introduction to the Project

LARC contraceptives are more effective forms of birth control than M/M methods and are safe for women of all ages to use. LARC methods could prevent more cases of teenage pregnancy for young women at risk and who are sexually active. However, M/M methods are more widely used and may be perceived by the public to be safer since they don't require surgical insertions of long-term devices into the body. Young women with inadequate knowledge about sex education and reproductive health may not have enough information about the different available contraceptive methods and as a result, be unable to make an informed decision on the contraceptive type that's right for them.

The primary research question being explored in this project is if adolescents (ages 15-20) use long-acting reversible methods of contraception (LARC) at a lower rate than older women (ages 21-44) and is that trend consistent throughout the three year period?







<br>




```{r read-data, echo=FALSE, cache=TRUE}
if (!file.exists("contraceptive_use.csv")) {
download.file("https://data.chhs.ca.gov/dataset/c2698502-d276-4e55-9057-8153e39d21b1/resource/12a73f54-dcf4-4e38-843c-e988385be69b/download/ofp-ccw-by-race-ethn_contra-type_age-group_14-16.csv", "contraceptive_use.csv", method="libcurl", timeout = 60) 
}
contra <- data.table::fread("contraceptive_use.csv") 
```

```{r data prep, echo=FALSE}
#remove the '%' symbol in the dataset
contra$`Rate of Contraceptive Use` <- stringr::str_remove_all(contra$`Rate of Contraceptive Use`, "%")

#column was changed from a character variable to numeric
contra$`Rate of Contraceptive Use` <- as.numeric(contra$`Rate of Contraceptive Use`)

#average rate by Contraceptive Type, Age Group and Year
avg_contra <- contra[ , .(
    avg_rate = mean (`Rate of Contraceptive Use`)
  ), 
  by = .(`Contraceptive Type`, `Age Group`, Year)]

#average rate by Race
race_contra <- contra[ , .(
    avg = mean (`Rate of Contraceptive Use`)
  ),
  by = `Race / Ethnicity`]
```



```{r plot1, echo=FALSE, class.source="code-r-small"}
p1_bar <- avg_contra %>% 
  ggplot()+
  geom_col(mapping=aes(x= Year, y= avg_rate, fill = `Age Group`), position = "dodge") +
    scale_fill_brewer(palette = "Set1") +
   labs( y = "Rate of Contraceptive Use", title = "Bar Graphs of Contraceptive Use from 2014-2016") +
  facet_wrap( ~ `Contraceptive Type`, nrow=2)
ggplotly_p1bar <- ggplotly(p1_bar)
```

```{r table1, echo=FALSE, class.source="code-r-small"}
contra <- as.data.frame(contra)
table1 <- contra %>% 
  select(Year,`Age Group`,`Race / Ethnicity`,`Contraceptive Type`,`Rate of Contraceptive Use`)
```

```{r plot2, echo=FALSE, class.source="code-r-small"}
p2_box <- contra %>% 
  ggplot()+
  geom_boxplot(mapping=aes(x=`Contraceptive Type`, y=`Rate of Contraceptive Use`, fill = `Contraceptive Type`)) +
  scale_fill_brewer(palette = "Set1") +
   labs(title = "Boxplots of Contraceptive Use by Year and Age Group") + 
    facet_wrap(Year ~ `Age Group`, nrow=3)
ggplotly_p2box <- ggplotly(p2_box)
```

```{r plot3, echo=FALSE, class.source="code-r-small"}
p3_bar <- race_contra %>% 
  ggplot()+
  geom_col(mapping=aes(x= `Race / Ethnicity`, y= avg, fill = `Race / Ethnicity`), position = "dodge") +
    scale_fill_brewer(palette = "Dark2") +
   labs( y = " Average Rate of Contraceptive Use", title = "Bar Graph of Rate of Contraceptive Use by Race") 
ggplotly_p3_bar <- ggplotly(p3_bar)
```







<br>









# Interactive Visuals {.tabset}

### Plot 1

```{r echo=FALSE}
ggplotly_p1bar
```

These bar graphs depict the rate of contraceptive use by age group and contraceptive type throughout the 3 year period. Most/Moderately Effective methods are used by both age groups almost 4 times more frequently than LARC methods. The trend between age groups is not consistent throughout the years. In 2015 and 2016, older women use LARC methods at a slightly higher rate than the younger age group. The only seemingly significant difference in contraceptive type between age groups is in 2014 for M/M contraception. 



### Table 1

```{r echo=FALSE}
datatable(table1)
```
This is an interactive data table displaying the rates of contraceptive type by year, age group, race/ethnicity and contraceptive type. You can utilize the search bar to display certain values or use the arrows at the top of the columns to reorder the values in each columns. 





### Plot 2

```{r echo=FALSE}
ggplotly_p2box
```

These interactive boxplots depict the distribution of contraceptive use rates by year and age group, with the contraceptive types shown side by side. Similar to the average rates shown on the bar graphs, these boxplots confirm that the median LARC contraceptive use rates for younger women are only higher than the rates for older women in 2014, but not the next two years. 



### PLot 3

```{r echo=FALSE}
ggplotly_p3_bar
```

This distribution shows the variability in contraceptive rates across the six ethnic groups. This bar graph indicates that women of Black and Hispanic race use contraception at a lower rate than women of other races. This observation should be considered for future research. There could potentially be lack of access, resources, health education for other barriers or these racial groups. Furthermore, the general variability in rates across ethnic groups could suggest that there are more underlying factors that affect women's access to contraception or openness to use it, such as cultural or religious values.




## {-}


# Conclusion
Although statistical analyses should be conducted in order to determine a more accurate conclusion, we can look at the figures and table to reach a preliminary conclusion. In reference to the formulated research question, younger women do indeed use LARC contraceptive methods at a lower rate than older women, but this was not true for all 3 years, 2014-2016. In 2014, younger women actually had higher rates for LARC contraceptive use. More research should be conducted as to why there was a change between 2014 and 2015 and if the trend has held the same past 2016. Introducing LARC contraceptive methods to younger women could reduce unwanted pregnancies and offer alternative, long-term contraception options.





